---
title: "Automatic vs Manual Gear"
author: "Sandro Vicente"
date: "Wednesday, June 17, 2015"
output: html_document
---

#### Overview

Overview Text

#### Analysis

```{r message=FALSE}
data(mtcars)
```

* Did the student interpret the coefficients correctly?

* Did the student fit multiple models and detail their strategy for model selection?

* Did the student answer the questions of interest or detail why the question(s) is (are) not answerable?

* Did the student do a residual plot and some diagnostics?
    
* Did the student quantify the uncertainty in their conclusions and/or perform an inference correctly?

* Was the report brief (about 2 pages long) for the main body of the report and no longer than 5 with supporting appendix of figures?


* Did the student do some exploratory data analyses?

We initially plot the variables with respect to mpg and transmission type in appendix plot 1.

Plot mpg X am shows a trend to higher mpg for manual transmission (red).

We can examine it using T-Test. For that we assume that the measurements of mpg and its gear type are independent and identically distributed.

```{r}
mpg_aut <- mtcars[mtcars$am==0, "mpg"]
mpg_man <- mtcars[mtcars$am==1, "mpg"]

t.test(mpg_man, mpg_aut, paired=F)
```

It shows the average mpg tends to be higher for manual transmission with 95% of confidence.

A basic linear model involving only mpg and transmission (am) confirms it.

```{r}
lm0 <- lm(mpg ~ factor(am)-1, data=mtcars)
summary(lm0)$coefficient
```

The coeficient for 1 (manual) is higher.

We check other variables against transmission type in the model. Both confounding and interactions are considered.

```{r}
vars <- c("cyl", "disp", "hp", "drat", "wt", "qsec", "gear", "carb" )
# Confounding
for (v in vars) {
    a<-anova(lm(as.formula(paste("mpg ", v, sep="~")), data=mtcars),
             lm(as.formula(paste("mpg ~ factor(am) ", v, sep="+")), data=mtcars))
    if (a$Pr[2] < 0.05) {
        print(v); print(a$Pr[2])
    }
}
# Interaction
for (v in vars) {
    a<-anova(lm(as.formula(paste("mpg ", v, sep="~")), data=mtcars),
             lm(as.formula(paste("mpg ~ factor(am) ", v, sep=":")), data=mtcars))
    if (a$Pr[2] < 0.05) {
        print(v); print(a$Pr[2])
    }
}
```

The variables "hp", "qsec" and "carb" are significantly impacted (both methods) by transmission type, with p-value lower that 0.05. They are considered for a refined modelling.

```{r}
fit_hp <- lm(mpg ~ factor(am) + hp -1, data=mtcars)
fit_qsec <- lm(mpg ~ factor(am) + qsec -1, data=mtcars)
fit_carb <- lm(mpg ~ factor(am) + factor(carb) -1, data=mtcars)

fit_hp_qsec <- update(fit_hp, mpg ~ factor(am) + hp * qsec -1)
anova(fit_hp, fit_hp_qsec) 
summary(fit_hp_qsec)$coefficients ## no good p-values

fit_hp_carb <- update(fit_hp, mpg ~ factor(am) + hp * carb -1)
anova(fit_hp, fit_hp_carb) # best p-values
summary(fit_hp_carb)$coefficients # best fit

fit_hp_qsec_carb <- update(fit_hp_carb, mpg ~ factor(am) + hp * qsec * carb -1)
anova(fit_hp_carb, fit_hp_qsec_carb) # no sig influence
summary(fit_hp_qsec_carb)$coefficients # no good p-values
```

The model associating mpg to am, hp and carb is the one with best p-values. The analysis of variance also show best results for this model.

#### Appedix

```{r }
par(mfrow=c(3,3),mar=c(4,4,1,0), oma=c(0,0,2,0))

with(mtcars, plot(disp, mpg, pch=19, col=am+1))
with(mtcars, plot(hp, mpg, pch=19, col=am+1)) # good
with(mtcars, plot(drat, mpg, pch=19, col=am+1))
with(mtcars, plot(wt, mpg, pch=19, col=am+1))
with(mtcars, plot(qsec, mpg, pch=19, col=am+1)) # good
with(mtcars, plot(gear, mpg, pch=19, col=am+1))
with(mtcars, plot(carb, mpg, pch=19, col=am+1))
with(mtcars, plot(cyl, mpg, pch=19, col=am+1))
with(mtcars, plot(am, mpg, pch=19, col=am+1))

legend("top", pch = 19, col = c(0,1)+1, legend = c("Auto", "Man"))
mtext("Plot 1 - mpg and transmission X other variables",side=3, line=1, outer=TRUE)

par(mfrow=c(2,2), mar=c(5,4,1,0), oma=c(0,0,2,0))
plot(fit_hp_carb) 
mtext("Plot 2 - Analysis of diagnostic values for fit_hp_carb",side=3, line=1, outer=TRUE)

```