---
title: "Automatic vs Manual Gear"
author: "Sandro Vicente"
output: html_document
---

#### Overview

Overview Text

#### Analysis

```{r message=FALSE, warning=FALSE}
library(car)
data(mtcars)
```

We initially plot the variables with respect to mpg and transmission type in appendix plot 1.

Plot mpg X am shows a trend to higher mpg for manual transmission (red).

We can examine it using T-Test. For that we assume that the measurements of mpg and its gear type are independent and identically distributed.

```{r}
mpg_aut <- mtcars[mtcars$am==0, "mpg"]
mpg_man <- mtcars[mtcars$am==1, "mpg"]
t.test(mpg_man, mpg_aut, paired=F)$conf.int
```

It shows the average mpg tends to be higher for manual transmission with 95% of confidence.

A basic linear model involving only mpg and transmission (am) confirms this.

```{r}
lm0 <- lm(mpg ~ factor(am)-1, data=mtcars)
summary(lm0)$coefficient
```

However a model involving only these variables may be innacurate. We check other variables against transmission type in the model. Both confounding and interactions are considered.

```{r}
vars <- c("cyl", "disp", "hp", "drat", "wt", "qsec", "gear", "carb" )
for (v in vars) {
    # confounding
    cf<-anova(lm(as.formula(paste("mpg ", v, sep="~")), data=mtcars),
             lm(as.formula(paste("mpg ~ factor(am) ", v, sep="+")), data=mtcars))
    # interaction only
    i<-anova(lm(as.formula(paste("mpg ", v, sep="~")), data=mtcars),
             lm(as.formula(paste("mpg ~ factor(am) ", v, sep=":")), data=mtcars))

    if (cf$Pr[2] < 0.05 && i$Pr[2] < 0.05) {
        print(v); print(cf$Pr[2]); print(i$Pr[2])
    }
}
```

The variables "hp", "qsec" and "carb" are significantly impacted (both methods) by transmission type, with p-value lower that 0.05. Therefore, they are considered for a refined modelling.

```{r}
# base models
fit_hp <- lm(mpg ~ factor(am) + hp -1, data=mtcars)
fit_qsec <- lm(mpg ~ factor(am) + qsec -1, data=mtcars)
fit_carb <- lm(mpg ~ factor(am) + factor(carb) -1, data=mtcars)

# function to show coeficients within specified confidence levels
sig_coefficients <- function(x, err) { s <- summary(x)$coefficients;s[s[,4]<err,]}

# model 1 - am, hp, qsec
fit_hp_qsec <- update(fit_hp, mpg ~ factor(am) + hp * qsec -1)
anova(fit_hp, fit_hp_qsec) 
sig_coefficients(fit_hp_qsec, 0.05) ## not many good p-values, only 'qsec'

# model 2 - am, hp, carb
fit_hp_carb <- update(fit_hp, mpg ~ factor(am) + hp * carb -1)
anova(fit_hp, fit_hp_carb) # best p-values
sig_coefficients(fit_hp_carb, 0.05) # best fit

# model 3 - am, hp, qsec, carb
fit_hp_qsec_carb <- update(fit_hp_carb, mpg ~ factor(am) + hp * qsec * carb -1)
anova(fit_hp_carb, fit_hp_qsec_carb) # no sig influence
sig_coefficients(fit_hp_qsec_carb, 0.05) # no good p-values
```

Model 2 associating mpg to am, hp and carb is the one with all coeficients have p-values under 0.05. 

Plot 2 in the appendix shows the diagnostic analysis for this model. No specific pattern is observed in the residuals charts. Normal Q-Q shows reasonable consistence between quantiles and residuals. Leverage chart shows strong influence of one one data point (Maserati Bora).

The coeficients are still consistent with the first test involving only transmission. Based on the data in mtcars, manual transmission is associated to higher mpg with 95% of confidence.  

The model 2 also shows that automatic transimssion corresponds to an increase of 4.73  +/- 2.56 in mpg. 

#### Appedix

```{r }
par(mfrow=c(3,3),mar=c(4,4,1,0), oma=c(0,0,2,0))

with(mtcars, plot(disp, mpg, pch=19, col=am+1))
with(mtcars, plot(hp, mpg, pch=19, col=am+1)) # good
with(mtcars, plot(drat, mpg, pch=19, col=am+1))
with(mtcars, plot(wt, mpg, pch=19, col=am+1))
with(mtcars, plot(qsec, mpg, pch=19, col=am+1)) # good
with(mtcars, plot(gear, mpg, pch=19, col=am+1))
with(mtcars, plot(carb, mpg, pch=19, col=am+1))
with(mtcars, plot(cyl, mpg, pch=19, col=am+1))
with(mtcars, plot(am, mpg, pch=19, col=am+1))

legend("top", pch = 19, col = c(0,1)+1, legend = c("Auto", "Man"))
mtext("Plot 1 - mpg and transmission X other variables",side=3, line=1, outer=TRUE)

par(mfrow=c(2,2), mar=c(5,4,1,0), oma=c(0,0,2,0))
plot(fit_hp_carb) 
mtext("Plot 2 - Analysis of diagnostic values for fit_hp_carb",side=3, line=1, outer=TRUE)

```