---
title: "Statistical Inference - Analysis of Exponential Dsistribution"
author: "Sandro Vicente"
date: "Saturday, May 16, 2015"
output: html_document
---
## Overview

 In a few (2-3) sentences explain what is going to be reported on.

## Simulation Part

In this part, we investigate the values from a exponential distribution, extracting means of 40 values of the distribution and observing how the observed values of the mean and standard errors adhere to the theretical values. We also investigate how distribution of means correspond to a normal distribution.

### Generating Data

Parameters used for the experiment

```{r}
nosim <- 1000 # number simulations
n <- 40
lambda <- 0.2 
```

The idea is to obtain `r nosim` means of `r n` values from the exponential distribution with lambda = `r lambda`. 
Therefore we generate a total of `r nosim * n` values from the exponential distribution, which we call *raw data*.

We can calculate the theoretical mean and standard deviations, which for exponential distributions, are based on lambda: 

```{r}
m.dist <- 1/lambda
sd.dist <- 1/lambda

m.dist # theoretical mean for the exponential distribution 
sd.dist # theoretical standard error for the exponential distribution
```

And generate the raw data using the exponential distribution. 
We also can calculate the mean and standard error for the raw data.

```{r}
ALL <- rexp(nosim*n, lambda) # all raw data

m.all <- mean(ALL)
sd.all <- sd(ALL)

m.all # mean for the raw values obtained
sd.all # standard error for the raw values obtained
```

We can see that the theoretical mean and standard deviations match the values for the raw data.

The Figure 1 shows a plot of the density of values from the raw data, which follows a typical exponential distribution curve.
The mean (marked in red) is the "centre of mass" of the curve.

```{r}
library(ggplot2)

DF.ALL <- data.frame(values=ALL)
ggplot(DF.ALL, aes(x=values)) + geom_histogram(binwidth=.1, aes(y= ..density..)) + xlim(0, 35) + xlab("All values") + ggtitle("Figure 1 - Density of raw samples") + geom_vline(aes(xintercept=5), colour="red")
```

Now we can calculate the mean of samples of `r n` values from the raw data. 

```{r}
MEAN <- matrix(ALL, nosim)   # split in blocks of 'n' samples
means <- apply(MEAN, 1, mean) # calculate the mean of each block
```


And calculate the mean and standard error for the means of 'n' samples
```{r}

m.means <- mean(means) 
sd.means <- sd(means) 

sd.n <- sd.all/sqrt(n) 

m.means  # mean of means of 'n' samples
sd.means # sd calculated from the means of 'n' samples
sd.n # sd based on original sd, corrected for mean of 'n' samples
```
Above, for the standard errors, calculating it directly from the means or using the value from the raw standard error divided by square root of `r n` result in similar values.

These results are compatible with the theoretical value for standard error of a sample of size 'n'=`r n` :
```{r}
sd.dist / sqrt(n) 
```

When we plot the distribution of *means of samples* of 'n' = `r n`, we obtain a curve that resambles a Gaussian, shown in figure 2. 
The red line shows the distribution's mean `r m.means`, which matches the area with highest density in the distribution. The blue line shows the theoretical mean `r m.dist`.

```{r}
DF <- data.frame(means=means)

ggplot(DF.ALL, aes(x=means)) + geom_histogram(binwidth=.1, aes(y= ..density..)) + ggtitle("Figure 2 - Density of means of 'n' samples") + geom_vline(aes(xintercept=m.means), colour="red") + geom_vline(aes(xintercept=m.dist), colour="blue")

```


### Analysis of the results

We can compare the distribution of means of samples agaist a normal distribution with the same mean and standard error. This is shown in the figure 3.

```{r}

ggplot(DF, aes(x=means)) + geom_histogram(alpha=0.3,binwidth=.1, aes(y= ..density..)) + stat_function(fun=dnorm, args=list(mean=m.means, sd=sd.means)) + ggtitle("Figure 3 - Means of 'n' samples against normal distribution ") 
```

This confirms the *Central Limit Theorem*, which states that the distribution of averages of iid variables becomes that of a standard normal as the sample size increases. 
In the case of this experiment, we used an sample of `r nosim` means of `r n` samples originated from an exponential distribution of lambda `r lambda`. The distribution of the means of 'n' values yielded results consistent with the ones of a normal distribution.
