---
title: "Analysis of vitamin C intake against teeth growth"
author: "Sandro Vicente"
output: html_document
---

#### Overview

In this report we analyse the influence of Vitamin C against odontoblasts (teeth) growth in Guinea Pigs. 
We compare both dosage and supplement types against the observed average teeth lengths, using both Normal distribution and Student's distribution. 

#### Data and Analysis

A first exploratory plot (figure 1) shows that lenght tend to increase with the dose. 
OJ supplement seems to have more influence in growth for doses of 0.5mg and 1mg.

```{r fig.width=6, fig.height=4} 
library(datasets)
# We can convert doses and supplement types into factors to perform further analysis.
tg <- ToothGrowth
tg$dose <- factor(tg$dose)
tg$supp <- factor(tg$supp)
confidence <- 0.95 
table(tg[,c("supp", "dose")])
```

We have the same number of observations for each combination of doses (`r levels(tg$dose)`) and and supplements (`r levels(tg$supp)`). All samples are assumed to be IID.

We can initially explore how the type of supplement alone affects variations in length. 
For all analysis we use the confidence interval of `r confidence`.

```{r}
calc_interval <- function(x, conf) {
    quantile <- 1-(1-conf)/2
    n <- length(x); sdn <- sd(x)/sqrt(n); m <- mean(x)
    list(quantile=quantile, n=n, sd=sdn, m=m, interval=(m+c(-1,1)*qnorm(quantile)*sdn) )
} 
```

```{r}
data.supp <- data.frame()
for (supp in levels(tg$supp)) {
    x=tg[tg$supp==supp,"len"]
    data.supp <- rbind(data.supp, data.frame(x=x, type=supp))
    l <- calc_interval(x, confidence)
    print(sprintf("Supplement: %s, mean: %f, sd: %f, size: %d, range:[%f, %f]", supp, l$m, l$sd, l$n, l$interval[1], l$interval[2]))
}
```

Both invervals overlap, therefore, following the normal distribution the supplement type alone cannot significantly determine the average teeth length of the population with a confidence of `r confidence*100`%.

Since the sample sizes are relatively small, we should use the T distribution to test the samples regarding supplement types. We assume that all observations are statistically independent.

```{r}
t.test(x ~ type, data=data.supp)$conf
```

Zero is in the confidence interval, which confirms that the supplement type cannot determine the teeth lenght.

It remains to be seen wheather the supplement type, for a given dosage, has impact on the average size of the teeth in a population.  
For that, we drill down into groups of observations for each supplement type and dosage.

```{r}
data.type <- data.frame()  # reset dataframe containg data per type (i.e. per dose and supp)
norm.type <- data.frame()  # reset dataframe containg summarized for normal distribution per type (i.e. per dose and supp)
for (dose in levels(tg$dose)) {
    for (supp in levels(tg$supp)) {
        x <- tg[tg$supp==supp & tg$dose==dose,"len"] # filter supp and dose
        l <- calc_interval(x, confidence)
        data.type <- rbind(data.type, data.frame(x=x, type=paste(supp,dose,sep="-"))) # type defined as composition supp "-" 
        norm.type <- rbind(norm.type, data.frame(m=l$m, sd=l$sd, type=paste(supp,dose,sep="-"))) # type defined as composition supp "-" dose
        print(sprintf("Supplement:%s, Dosage: %s mg, mean: %f, sd: %f, size: %d, range:[%f, %f]", supp, dose, l$m, l$sd, l$n, l$interval[1], l$interval[2]))
    }
}
```

From the values obtained above we can see that the supplement type has more influence in the average lenght for lower dosages, especially for 0.5 mg.
For this dosage, the OJ correspond to a greater average teeth leght, compared to VC.

Figure 2 shows the average length distributions for each group and corresponding gaussian curves.

The samples for these groups are the smallest and we can observe a number of overlaps on the normal distributions. It is important to confirm the results with T tests. The results are shown on table 1

Using Student's T distribution, the intervals show that the lenghts can be significantly (`r confidence*100`%) mapped into each of the groups with only the exceptions of groups OJ-1 agaist VC-2 and OJ-2 against VC-2, where the confidence interval of the differences in length include zero. 

These results confirm that higher dosages correspond to greather mean teeth lengtht, and that at lower dosages, the supplement has stronger influence on the average teeth length. Specifically, the OJ supplement is more effective in increasing the length.
OJ for 1mg is not distinguishable from VC for 2mg, and for the dosage of 2mg, both OJ and VC supplements have similar infuence.


#### Appendix: Charts and Tables

```{r fig.width=8, fig.height=4, message=FALSE}
library(ggplot2) # auxiliry libraries added
library(plyr)

# Calculate normal values for reference 
len.min <- min(tg$len) # minimum length
len.max <- max(tg$len) # maximum legth
len.x <- seq(len.min, len.max, length=100) # range of lengths from min to max 

# function to calculate normal densities given data frame containg "type", "m" mean and "sd" sample's standard deviation
calc_normal <- function(nt) {
    ddply(norm.type, "type", function(df) {
        data.frame( 
            predicted = len.x,
            density = dnorm(len.x, mean=df$m, sd=df$sd)
        )
    })
}

## shows clusters
ggplot(data=ToothGrowth, aes(x=len, y=dose, group=supp, colour=supp)) +
    geom_point() + ggtitle("Fig. 1 - Distribution of length per supplement type and dose") 

# calculate normal distribution for types based on dose and supp
normaldens <- calc_normal(norm.type)

# plot histogram and corresponding normal distributions althogether
ggplot(data.type, aes(x=x, fill=type)) +  geom_histogram(alpha = 0.5, aes(y = ..density..), position = 'identity') +
    geom_line(data=normaldens, aes(x=predicted, y=density, color=type))+ 
    xlab("Length") + ggtitle("Fig.2 - Length and avg. distribution / dosage and supplement (overlay)") 
```

**Table 1 - Student's confidence intervals for each group of dosage and supplement**
``` {r}

apply(combn(levels(data.type$type),2, simplify=T), 2, function(t) {
    conf <- t.test(x ~ type, data=subset(data.type, type %in% t))$conf
    c(t, conf, ifelse(conf[1]*conf[2]>0,"OK","UNCONFIRMED"))
})
    
```
