---
title: "Analysis of vitamin C intake against tooth growth"
author: "Sandro Vicente"
date: "Sunday, May 17, 2015"
output: html_document
---

### Overview

to do...

### Obtaining Data

Data is obtained from standard R datasets.

```{r} 
library(datasets)

library(ggplot2)
## shows clusters
ggplot(data=ToothGrowth, aes(x=len, y=dose, group=supp, colour=supp)) +
    geom_point() + ggtitle("Figure 1 - Distribution of length per supplement type and dose") 

```

A first exploratory plot (figure 1) shows that lenght tend to increase with the dose. 
OJ supplement seems to have more influence in growth for doses of 0.5mg and 1mg.

We can convert doses and supplement types into factors to perform further analysis.

```{r}
tg <- ToothGrowth
tg$dose <- factor(tg$dose)
tg$supp <- factor(tg$supp)

summary(tg)

table(tg[,c("supp", "dose")])
```
We have the same number of observations for each combination of doses (`r levels(tg$dose)`) and and supplements (`r levels(tg$supp)`).

We can initially explore how the type of supplement alone affects variations in length.

For that, a function is defined to take a number of samples and estimate the mean, standard error and the limits of values for a confidence interval using the Central Limit Theorem.

*Here we assume all samples are IID*
```{r}
calc_interval <- function(x, conf) {
    quantile <- 1-(1-conf)/2
    n <- length(x)
    sd <- sd(x)/sqrt(n)
    m <- mean(x)
    interval <- (m+c(-1,1)*qnorm(quantile)*sd)    
    list(quantile=quantile, n=n, sd=sd, m=m, interval=interval)
} 

confidence <- 0.95 
```

Analysis over different supplement types. 
For all analysis we use the confidence interval of `r confidence`.
```{r}
for (supp in levels(tg$supp)) {
    x <- tg[tg$supp==supp,"len"]
    l <- calc_interval(x, confidence)
    print(sprintf("Supplement: %s, mean: %f, sd: %f, size: %d, range:[%f, %f]", supp, l$m, l$sd, l$n, l$interval[1], l$interval[2]))
}
```
For the given confidence, we can see that both invervals overlap, so the supplement type alone is not significat to correspond the length.

Analysis over different doses
```{r}
for (dose in levels(tg$dose)) {
    x <- tg[tg$dose==dose,"len"]
    l <- calc_interval(x, confidence)
    print(sprintf("Dosage: %s, mean: %f, sd: %f, size: %d, range:[%f, %f]", dose, l$m, l$sd, l$n, l$interval[1], l$interval[2]))
}
```

Drill down into supplement types and doses
```{r}
for (dose in levels(tg$dose)) {
    for (supp in levels(tg$supp)) {
        print(c(dose,supp))

        dose.supp <- tg[tg$supp==supp & tg$dose==dose,"len"]
        print(calc_interval(dose.supp, confidence))
    }
}
```
